Quantifying passive kelp rafting phenomena and connectivity of nearshore communities model around the coast of South Africa:Development of a 3D spectral numerical coastal-flow

### 2020-06-29 thesis_chapter_1.tex
supporting files

---
title: "Quantifying passive kelp rafting phenomena and connectivity of nearshore communities around the coast of South Africa:Development of a 3D spectral numerical coastal-flow model"
author: 
- RM Coppin^[University of the Western Cape]
- C Rautenbach^[South African Weather Service]
- AJ Smit^[University of the Western Cape]
output:
  html_document:
    df_print: paged
    toc: yes
    theme:  simplex
  word_document:
    fig_caption: yes
    reference_docx: chapter_1_template.docx
    toc: yes
bibliography: references.bib
---

```{r include=FALSE}
library(png)
library(grid)
```

# Summary

Marine ecosystems are maintained by a variety of complex interactions between abiotic and biotic variables such as temperature, wave exposure, pH, competition, and processes such as top-down and bottom-up control, predator-prey relationships and phenology [@doney2011; @Harley2012; @Burrows2011; @McGowan1998]. These abiotic and biotic variables, the interactions between them, and the various ecological processes, ultimately determine the community composition and ecological functioning of all ecosystems [@Guimaraes1996; @Harley2006; @Poloczanska2013; @Jennings2010; @Polovina2005; @Johnson2011; @Krumhansl2016; @Dayton1999]. Climate directly and indirectly affects the way in which abiotic and biotic variables interact, but is often compounded by other impacts such as habitat destruction, pollution, and over-fishing [@Blamey2012; @Blamey2015]. Temperature and wave exposure have been recognised as important variables with regards to climate-driven changes within the ocean (Guimaraes & Coutinho 1996; McGowan et al. 1998; McQuaid & Branch 1984; Laufkötter et al. 2015; Belkin 2009; Filbee-Dexter et al. 2018; Miller et al. 2011; Smale et al. 2011; Smale & Moore 2017). In order to persist and survive within variable and changing environments, organisms must either migrate, adapt, or die.

Seaweeds are sessile organisms that are unable to migrate to new areas when local environmental conditions become unsuitable and therefore are forced to adapt to new conditions in order to avoid expiration (Dayton 1985; Tegner et al. 1997). The main form of mortality for seaweeds is through mechanical dislodgement by wave action (Schiel et al. 2006; Seymour et al. 1989; Thomsen et al. 2004; Cavanaugh et al. 2011; Graham et al. 1997; Demes et al. 2013; Edwards & Estes 2006). Seaweeds, particularly brown seaweeds, are able to undergo rapid morphological adaptation to the hydrodynamic environment (Miller et al. 2011; Friedland & Denny 1995; Wing et al. 2007; Wernberg & Thomsen 2005; Fowler-Walker et al. 2006; Dudgeon & Johnson 1992; Blanchette et al. 2002). This allows seaweeds to reduce mortality through mechanical dislodgement by inducing morphology which reduces overall drag. Seaweeds that are unable to avoid mechanical dislodgement either raft out to sea or wash up onto beaches. Not all beach-cast kelp may have originated from a nearby kelp population and may have originated from other sites or regions of the coast through rafting and ocean currents (Malm et al. 2004; Emery & Tschudy 1941; Filbee-Dexter & Scheibling 2012; Hobday 2000). Therefore, because kelp morphology is specific to its local environment the morphological features may be able to indicate, within a certain amount of probability, what site or region it most likely originated from. In other words, beach-cast kelp may be used as proxy for investigating the flow of coastal currents. Using kelp as a proxy for determining its original location will be calibrated by means of a hydrodynamic model which will be designed from already existing SWAN and Delft3D models. This combined approach will allow investigation into flow regimes around the west and south-west coasts of South Africa and the role they play in subtidal ecology. Furthermore, the model could be used to determine transport of plastic pollution around the coast. For instance, micro-plastics are recognised as a threat to marine life (Seltenrich 2015; Azzarello & Van Vleet 1987; Wright et al. 2013; Derraik 2002; Andrady 2011; Fendall & Sewell 2009; Ivar Do Sul & Costa 2014; O ’donoghue & Marshall n.d.; Ballent et al. 2012), however very little is known about how microplastics may be transported along the coast.

The increased use of plastic in society over the past half century has resulted in large amounts of plastic litter in both the marine and terrestrial environment (Andrady 2011; Wright et al. 2013; Cole et al. 2011). The problems associated with large plastic debris have received attention for many decades, whereas those connected to marine micro-plastics comparatively received very little attention. However, today it has become a prioritized area among political organizations, agencies and NGOs around the world. The micro-plastic debris present in the ocean are derived from marine and terrestrial sources (Derraik 2002; Wright et al. 2013; Cole et al. 2011; Seltenrich 2015), however there is little understanding of how microplastics may be distributed with ocean currents. Therefore, the coupling the adaptability of kelp morphology and the ability to simulate hydrodynamic processes can greatly improve our understanding of transport pathways and likely locations of accumulation. This in turn may inform management decisions with regards to eliminating and managing marine pollution in South Africa.

# Background

@@

Advances in numerical modelling has gained much traction in recent years and has been applied in a variety of ways with regards to ecological studies. For example, a study by Wang and Xia (2009) used the Delft3D-Flow model to assess the hydraulic suitability of a stream as a spawning ground for the Chinese Sturgeon (Acipenser sinensis) in the Yangtze River. The authors calculated the horizontal mean velocity which was used to assess the hydraulic environment of spawning ground. The flow field state was determined through model simulation and field-measured data used to validate the model. The results added to existing scientific database for spawning ground hydraulic environmental protection. Different numerical models can often be integrated to model across ecosystem levels. For example a study by Leon et al. (2003) used integrated physical (Delft3D hydrological model) and bio-chemical (Agricultural Non-point Source model) processes models to investigate the possible impact on the Lake Malawi water quality due to management actions performed at the watershed level.

Since wave energy is an important driver in marine ecosystems, particularly kelp, the advances in hydrodynamic modelling offer a new opportunity for multifactorial and quantitative approach to research in marine ecosystems. The Delft3D and SWAN models have been used successfully in previous studies regarding brine plume discharge, impacts of storms, effects of climate change on the hydrological environment etc. The models have not been designed for shallow environments (<6m) and therefore may not be suitable to model coastal hydrological environments. However these models may be adjusted to suit coastal waters if they are combined with a new numerical model which can be calibrated to suit these needs.

In recent years there has been growing attention on plastic pollution, particularly in the ocean. Plastic pollution can be in the form of macro- and microplastics. Microplastics are tiny plastic granules used as scrubbers in cosmetics and air-blasting, and small plastic fragments that originate from larger pieces of plastic known as macroplastics, while macroplastics…insert definition here… The potential harms of of plastic pollution in the marine environment was highlighted in the 1970’s and renewed interest has lead to research showing that plastic pollution in the ocean are widespread. Plastics may become bio-available to biota in the food-web which may cause problems with an organism's physiological functioning. Furthermore, the relatively large surface area and composition of microplastics provides an environment that is able of adhering to organic pollutants. In other words microplastics also act as a vector for transport and assimilation of organic pollutants.

@@

# Kelp environmental drivers

Important environmental drivers of kelp individuals and communities include light, substrata, salinity, sedimentation, nutrients, temperature and wave exposure. Although studies have investigated the effects of these important environmental drivers, the roles these factors play is often difficult to evaluate as such factors may never be fully independent of each other, i.e. environmental factors are to some extent are dependent on one another. Multifactorial studies have attempted to explain combined effects, however these studies are often limited to investigating combination of two or three environmental drivers as inclusion of too many factors can lead to results that are difficult to interpret. Environmental factors are highly variable on temporal and spatial scales and their effects may also be dependent on the life-stage of the organism, adding a further layer of complexity to investigations.

## Light

@@

## Depth

Depth does not affect kelp ecosystems directly, however a change in depth causes fluctuations or changes in other environmental variables such as water motion, light and temperature. Water motion also decreases with depth, and some kelps better suited to deeper environments (*L. pallida*) replace those in the shallows (*E. maxima*) (Dayton 1985; Gerard 1982). The increase in depth can lead to a decrease in sunlight penetration, with some species better adapted for low-light conditions than others, such as (*L. pallida*). Temperature may also change along a depth gradient due to a reduction in sunlight penetration (Dayton 1985; Gerard 1982). Therefore depth does not directly play a role in kelp functioning but may alter more influential factors such as light and water motion.

## Nutrients

@@

# The mechanisms of morphological adaptation
## Waves and macroalgae characteristics

In the wave swept nearshore one would expect organisms to reflect the harsh hydrodynamic environment by being streamlined, small and amoured. This is certainly the case for an array of fauna which often comprise of hard, rigid bodies that are held firmly in place to the substratum such as limpets and isopods......

# Ocean and coastal waves
## Introduction

Regardless of the location around the world, waves are a feature of any coastline and the marine environment, and are important manifestations of energy in the ocean. Waves are not the movement of water particles but rather the movement and propagation of energy through the ocean. The source of this energy can be formed locally, such as wind-driven waves, and or from distant locations in the ocean such as storms, know as *swell*. This energy is transferred from deeper water into shallower water where it plays a role in driving complex marine ecosystems as well as shaping the environments that they live in. Energy that is left-over after these processes is transferred into heat energy, and heats up the sand and rocks on which each wave hits. This energy can also be harnessed in simple ways such as a surfer catching a wave, or more complex ways such as capturing energy from the ocean environment for electricity production. However, waves are stochastic in nature and therefore the energy that propagates through the ocean is not always consistent. Therefore, it is often difficult to quantify and predict wave energy in the marine environment. Despite this fact, waves are gaining more recognition for the role they play in shaping coastlines, beaches and hence the communities and organisms that depend on these systems.

Waves manifest themselves in different ways, which is also dependent on the energy creating force, and can be classified into different categories. For instance, "chop" are produced by local winds, while “tsunamis”" are rare waves that are formed during a earthquakes or landslides and can be produced from thousands of kilometers away. Waves in the ocean are known as “swell” and are produced from far distant storms, while the most consistent form of waves that interacts with the coastlines, usually twice a day, are called “tides”. Tides are produced from differences in gravitational forces between the ocean, earth’s crust, sun and the moon. These classifications can be broken down further and will be investigated later in this chapter. The focus of this chapter will be understanding wave theory and how waves form and propagate into shallow-coastal waters. This will be essential to later chapters that aim to model wave energy and calibrate the produced model with biotic variables.

## Generating and restoring forces

Waves are formed due to the constant interaction between two forces, which are known as the "generating force" and the “restoration force”. The generating force is the force which pushes water from one layer up into the other. Layers in the ocean are formed through differences in temperature (thermocline) and salinity (pycnocline) which creates a boundry for wave energy to move along. In other words, waves occur along the boundaries formed in the ocean by various abiotic processes, and therefore density boundaries are essential to the propagation of energy through the ocean.

The generating force and restoring force occur along these boundaries, the generating force pushes water up across ocean boundaries while the restoring force pulls it back to where the boundary was originally, trying to restore the balance in energy. This "tug of war" between the two forces creates an oscillating motion between the boundary layers which acts as a point of disturbance, sending out energy in all directions. The disturbance energy will continue to manifest itself provided the tug and pull between the two forces is occurring. Once the generating force stops, and ultimately the energy from the point of disturbance, the waves dissipate and water restores to its original state. A simple example would be blowing on water in a cup. The blowing of air onto the surface of the water in the cup creates a point of disturbance that pushes the air boundary into the water boundary. The restoring force is the surface tension of the water that is maintained through hydrogen bonds between water molecules. Once blowing on the water in the cup has stopped, the ripples or “waves” in the cup begin to diminish in size until the surface tension returns to its original state. The hydrogen bonds between the water molecules are stronger than the force of gravity, and therefore the force of the surface tension returns the water to its original state. These waves are known as “capillary waves” and is essentially residual energy after the generating source has stopped. The restoring force can also be in the form of gravity and are known as “gravity waves”. Using the same example, if one blows too hard, the water boundary is pushed up into the air boundary, breaking the hydrogen bonds between water molecules which allows gravity to return the water to its original state. All waves in the ocean are either capillary waves or gravity waves, and their classification will be dependent on the restoring force involved.

@@

## Types of waves

There are a variety of waves that form in the ocean and all differ in terms of period or wavelength (??image??). The longest wave that can form in the ocean are known as trans-tidal waves, and are generated by fluctuations in magnetism between the Earth’s crust and atmosphere. The magnetic push and pull between the Moon and the Sun creates waves with a slightly shorter wavelength, known as tides. Their period and wavelength can also range from a few hours to more than a day, and from a few hundred to a thousand kilometers respectively. Storm surge tend to have a slightly shorter wavelength and period compared to tides. When low atmospheric pressure systems and high wind speeds in a storm it elevates the ocean surface, generating storm surge, which may cause flooding in coastal areas as it approaches the coastline. Tsunami’s are on the lower end of the scale, and are generated by earthquakes or submarine ‘landslides. Their random nature makes them difficult to predict and increase amplitude as they approach the coast. This can make them considerably large along coastal areas and often cause immense damage and loss of life.

## Measuring waves

Waves are often thought of as an elevation of the sea surface from a specific point over a period of time but this is obviously not he case. This is known as *surface elevation* and is the instantaneous elevation of the sea surface above a specific point in a time record (see figure \ref{fig:surface elevation figures}). Although surface elevation does not represent a wave, it can be used to create a wave profile. This is achieved by profiling the the surface elevation between two successive *downward zero-crossings* or *upwardward zero-crossings* (see figure \ref{fig:surface elevation figures}). Both zero-crossings are symmetrical and essentially the same statistically. However, in practice the downward zero-crossings are preferred as it takes steepness of a approaching wave into account (the front, see figure \ref{fig:surface elevation figures}) which is relevant to characterising breaking waves. It is important to note that surface elevation can be negative while a wave profile cannot.

```{r surface elevation figures, echo=FALSE, warning=FALSE, fig.align="center", fig.cap="\\label{fig:surface elevation figures}The definition of a wave in a time record of the surface elevation with downward zero-crossings (upper panel) and upward zero-crossings (lower panel)."}

surface_elevation <- readPNG("figures/surface_elevation_waves.png")
grid.raster(surface_elevation)
```

@@

```{r Hs_Tp_surface_elevation figure, echo=FALSE, warning=FALSE, fig.align="center", fig.cap="The definition of wave height and wave period in a time record of the surface elevation (the wave is defined with downwrd zero-crossings."}

Hs_Tp_elevation <- readPNG("figures/Hs_Tp_surface_elevation.png")
grid.raster(Hs_Tp_elevation)
```

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Quantifying passive kelp rafting phenomena and connectivity of nearshore communities model around the coast of South Africa:Development of a 3D spectral numerical coastal-flow},
  pdfauthor={RM Coppin; C Rautenbach; AJ Smit},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% https://github.com/rstudio/rmarkdown/issues/337
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

% https://github.com/rstudio/rmarkdown/pull/252
\usepackage{titling}
\setlength{\droptitle}{-2em}

\pretitle{\vspace{\droptitle}\centering\huge}
\posttitle{\par}

\preauthor{\centering\large\emph}
\postauthor{\par}

\predate{\centering\large\emph}
\postdate{\par}

\title{Quantifying passive kelp rafting phenomena and connectivity of nearshore
communities model around the coast of South Africa:Development of a 3D
spectral numerical coastal-flow}
\author{RM Coppin\footnote{University of the Western Cape} \and C Rautenbach\footnote{South African Weather Service} \and AJ Smit\footnote{University of the Western Cape}}
\date{}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{summary}{%
\section{Summary}\label{summary}}

Marine ecosystems are maintained by a variety of complex interactions
between abiotic and biotic variables such as temperature, wave exposure,
pH, competition, and processes such as top-down and bottom-up control,
predator-prey relationships and phenology (Doney et al. 2011; Harley et
al. 2012; Burrows et al. 2011; McGowan, Cayan, and Dorman 1998). These
abiotic and biotic variables, the interactions between them, and the
various ecological processes, ultimately determine the community
composition and ecological functioning of all ecosystems
({\textbf{???}}; Harley et al. 2006; Poloczanska et al. 2013; Jennings
and Brander 2010; Polovina 2005; Johnson et al. 2011; Krumhansl et al.
2016; Dayton et al. 1999). Climate directly and indirectly affects the
way in which abiotic and biotic variables interact, but is often
compounded by other impacts such as habitat destruction, pollution, and
over-fishing (Blamey and Branch 2012; Blamey et al. 2015). Temperature
and wave exposure have been recognised as important variables with
regards to climate-driven changes within the ocean (Guimaraes \&
Coutinho 1996; McGowan et al.~1998; McQuaid \& Branch 1984; Laufkötter
et al.~2015; Belkin 2009; Filbee-Dexter et al.~2018; Miller et al.~2011;
Smale et al.~2011; Smale \& Moore 2017). In order to persist and survive
within variable and changing environments, organisms must either
migrate, adapt, or die.

Seaweeds are sessile organisms that are unable to migrate to new areas
when local environmental conditions become unsuitable and therefore are
forced to adapt to new conditions in order to avoid expiration (Dayton
1985; Tegner et al.~1997). The main form of mortality for seaweeds is
through mechanical dislodgement by wave action (Schiel et al.~2006;
Seymour et al.~1989; Thomsen et al.~2004; Cavanaugh et al.~2011; Graham
et al.~1997; Demes et al.~2013; Edwards \& Estes 2006). Seaweeds,
particularly brown seaweeds, are able to undergo rapid morphological
adaptation to the hydrodynamic environment (Miller et al.~2011;
Friedland \& Denny 1995; Wing et al.~2007; Wernberg \& Thomsen 2005;
Fowler-Walker et al.~2006; Dudgeon \& Johnson 1992; Blanchette et
al.~2002). This allows seaweeds to reduce mortality through mechanical
dislodgement by inducing morphology which reduces overall drag. Seaweeds
that are unable to avoid mechanical dislodgement either raft out to sea
or wash up onto beaches. Not all beach-cast kelp may have originated
from a nearby kelp population and may have originated from other sites
or regions of the coast through rafting and ocean currents (Malm et
al.~2004; Emery \& Tschudy 1941; Filbee-Dexter \& Scheibling 2012;
Hobday 2000). Therefore, because kelp morphology is specific to its
local environment the morphological features may be able to indicate,
within a certain amount of probability, what site or region it most
likely originated from. In other words, beach-cast kelp may be used as
proxy for investigating the flow of coastal currents. Using kelp as a
proxy for determining its original location will be calibrated by means
of a hydrodynamic model which will be designed from already existing
SWAN and Delft3D models. This combined approach will allow investigation
into flow regimes around the west and south-west coasts of South Africa
and the role they play in subtidal ecology. Furthermore, the model could
be used to determine transport of plastic pollution around the coast.
For instance, micro-plastics are recognised as a threat to marine life
(Seltenrich 2015; Azzarello \& Van Vleet 1987; Wright et al.~2013;
Derraik 2002; Andrady 2011; Fendall \& Sewell 2009; Ivar Do Sul \& Costa
2014; O 'donoghue \& Marshall n.d.; Ballent et al.~2012), however very
little is known about how microplastics may be transported along the
coast.

The increased use of plastic in society over the past half century has
resulted in large amounts of plastic litter in both the marine and
terrestrial environment (Andrady 2011; Wright et al.~2013; Cole et
al.~2011). The problems associated with large plastic debris have
received attention for many decades, whereas those connected to marine
micro-plastics comparatively received very little attention. However,
today it has become a prioritized area among political organizations,
agencies and NGOs around the world. The micro-plastic debris present in
the ocean are derived from marine and terrestrial sources (Derraik 2002;
Wright et al.~2013; Cole et al.~2011; Seltenrich 2015), however there is
little understanding of how microplastics may be distributed with ocean
currents. Therefore, the coupling the adaptability of kelp morphology
and the ability to simulate hydrodynamic processes can greatly improve
our understanding of transport pathways and likely locations of
accumulation. This in turn may inform management decisions with regards
to eliminating and managing marine pollution in South Africa.

\hypertarget{background}{%
\section{Background}\label{background}}

@@

Advances in numerical modelling has gained much traction in recent years
and has been applied in a variety of ways with regards to ecological
studies. For example, a study by Wang and Xia (2009) used the
Delft3D-Flow model to assess the hydraulic suitability of a stream as a
spawning ground for the Chinese Sturgeon (Acipenser sinensis) in the
Yangtze River. The authors calculated the horizontal mean velocity which
was used to assess the hydraulic environment of spawning ground. The
flow field state was determined through model simulation and
field-measured data used to validate the model. The results added to
existing scientific database for spawning ground hydraulic environmental
protection. Different numerical models can often be integrated to model
across ecosystem levels. For example a study by Leon et al.~(2003) used
integrated physical (Delft3D hydrological model) and bio-chemical
(Agricultural Non-point Source model) processes models to investigate
the possible impact on the Lake Malawi water quality due to management
actions performed at the watershed level.

Since wave energy is an important driver in marine ecosystems,
particularly kelp, the advances in hydrodynamic modelling offer a new
opportunity for multifactorial and quantitative approach to research in
marine ecosystems. The Delft3D and SWAN models have been used
successfully in previous studies regarding brine plume discharge,
impacts of storms, effects of climate change on the hydrological
environment etc. The models have not been designed for shallow
environments (\textless6m) and therefore may not be suitable to model
coastal hydrological environments. However these models may be adjusted
to suit coastal waters if they are combined with a new numerical model
which can be calibrated to suit these needs.

In recent years there has been growing attention on plastic pollution,
particularly in the ocean. Plastic pollution can be in the form of
macro- and microplastics. Microplastics are tiny plastic granules used
as scrubbers in cosmetics and air-blasting, and small plastic fragments
that originate from larger pieces of plastic known as macroplastics,
while macroplastics\ldots insert definition here\ldots{} The potential
harms of of plastic pollution in the marine environment was highlighted
in the 1970's and renewed interest has lead to research showing that
plastic pollution in the ocean are widespread. Plastics may become
bio-available to biota in the food-web which may cause problems with an
organism's physiological functioning. Furthermore, the relatively large
surface area and composition of microplastics provides an environment
that is able of adhering to organic pollutants. In other words
microplastics also act as a vector for transport and assimilation of
organic pollutants.

@@

\hypertarget{kelp-environmental-drivers}{%
\section{Kelp environmental drivers}\label{kelp-environmental-drivers}}

Important environmental drivers of kelp individuals and communities
include light, substrata, salinity, sedimentation, nutrients,
temperature and wave exposure. Although studies have investigated the
effects of these important environmental drivers, the roles these
factors play is often difficult to evaluate as such factors may never be
fully independent of each other, i.e.~environmental factors are to some
extent are dependent on one another. Multifactorial studies have
attempted to explain combined effects, however these studies are often
limited to investigating combination of two or three environmental
drivers as inclusion of too many factors can lead to results that are
difficult to interpret. Environmental factors are highly variable on
temporal and spatial scales and their effects may also be dependent on
the life-stage of the organism, adding a further layer of complexity to
investigations.

@@

Depth does not affect kelp ecosystems directly, however a change in
depth causes fluctuations or changes in other environmental variables
such as water motion, light and temperature. Water motion also decreases
with depth, and some kelps better suited to deeper environments
(\emph{L. pallida}) replace those in the shallows (\emph{E. maxima})
(Dayton 1985; Gerard 1982). The increase in depth can lead to a decrease
in sunlight penetration, with some species better adapted for low-light
conditions than others, such as (\emph{L. pallida}). Temperature may
also change along a depth gradient due to a reduction in sunlight
penetration (Dayton 1985; Gerard 1982). Therefore depth does not
directly play a role in kelp functioning but may alter more influential
factors such as light and water motion.

@@

\hypertarget{the-mechanisms-of-morphological-adaptation}{%
\section{The mechanisms of morphological
adaptation}\label{the-mechanisms-of-morphological-adaptation}}

\hypertarget{waves-and-macroalgae-characteristics}{%
\subsection{Waves and macroalgae
characteristics}\label{waves-and-macroalgae-characteristics}}

In the wave swept nearshore one would expect organisms to reflect the
harsh hydrodynamic environment by being streamlined, small and amoured.
This is certainly the case for an array of fauna which often comprise of
hard, rigid bodies that are held firmly in place to the substratum such
as limpets and isopods\ldots\ldots{}

\hypertarget{ocean-and-coastal-waves}{%
\section{Ocean and coastal waves}\label{ocean-and-coastal-waves}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Regardless of the location around the world, waves are a feature of any
coastline and the marine environment, and are important manifestations
of energy in the ocean. Waves are not the movement of water particles
but rather the movement and propagation of energy through the ocean. The
source of this energy can be formed locally, such as wind-driven waves,
and or from distant locations in the ocean such as storms, know as
\emph{swell}. This energy is transferred from deeper water into
shallower water where it plays a role in driving complex marine
ecosystems as well as shaping the environments that they live in. Energy
that is left-over after these processes is transferred into heat energy,
and heats up the sand and rocks on which each wave hits. This energy can
also be harnessed in simple ways such as a surfer catching a wave, or
more complex ways such as capturing energy from the ocean environment
for electricity production. However, waves are stochastic in nature and
therefore the energy that propagates through the ocean is not always
consistent. Therefore, it is often difficult to quantify and predict
wave energy in the marine environment. Despite this fact, waves are
gaining more recognition for the role they play in shaping coastlines,
beaches and hence the communities and organisms that depend on these
systems.

Waves manifest themselves in different ways, which is also dependent on
the energy creating force, and can be classified into different
categories. For instance, ``chop'' are produced by local winds, while
``tsunamis''" are rare waves that are formed during a earthquakes or
landslides and can be produced from thousands of kilometers away. Waves
in the ocean are known as ``swell'' and are produced from far distant
storms, while the most consistent form of waves that interacts with the
coastlines, usually twice a day, are called ``tides''. Tides are
produced from differences in gravitational forces between the ocean,
earth's crust, sun and the moon. These classifications can be broken
down further and will be investigated later in this chapter. The focus
of this chapter will be understanding wave theory and how waves form and
propagate into shallow-coastal waters. This will be essential to later
chapters that aim to model wave energy and calibrate the produced model
with biotic variables.

\hypertarget{generating-and-restoring-forces}{%
\subsection{Generating and restoring
forces}\label{generating-and-restoring-forces}}

Waves are formed due to the constant interaction between two forces,
which are known as the ``generating force'' and the ``restoration
force''. The generating force is the force which pushes water from one
layer up into the other. Layers in the ocean are formed through
differences in temperature (thermocline) and salinity (pycnocline) which
creates a boundry for wave energy to move along. In other words, waves
occur along the boundaries formed in the ocean by various abiotic
processes, and therefore density boundaries are essential to the
propagation of energy through the ocean.

The generating force and restoring force occur along these boundaries,
the generating force pushes water up across ocean boundaries while the
restoring force pulls it back to where the boundary was originally,
trying to restore the balance in energy. This ``tug of war'' between the
two forces creates an oscillating motion between the boundary layers
which acts as a point of disturbance, sending out energy in all
directions. The disturbance energy will continue to manifest itself
provided the tug and pull between the two forces is occurring. Once the
generating force stops, and ultimately the energy from the point of
disturbance, the waves dissipate and water restores to its original
state. A simple example would be blowing on water in a cup. The blowing
of air onto the surface of the water in the cup creates a point of
disturbance that pushes the air boundary into the water boundary. The
restoring force is the surface tension of the water that is maintained
through hydrogen bonds between water molecules. Once blowing on the
water in the cup has stopped, the ripples or ``waves'' in the cup begin
to diminish in size until the surface tension returns to its original
state. The hydrogen bonds between the water molecules are stronger than
the force of gravity, and therefore the force of the surface tension
returns the water to its original state. These waves are known as
``capillary waves'' and is essentially residual energy after the
generating source has stopped. The restoring force can also be in the
form of gravity and are known as ``gravity waves''. Using the same
example, if one blows too hard, the water boundary is pushed up into the
air boundary, breaking the hydrogen bonds between water molecules which
allows gravity to return the water to its original state. All waves in
the ocean are either capillary waves or gravity waves, and their
classification will be dependent on the restoring force involved.

@@

\hypertarget{types-of-waves}{%
\subsection{Types of waves}\label{types-of-waves}}

There are a variety of waves that form in the ocean and all differ in
terms of period or wavelength (??image??). The longest wave that can
form in the ocean are known as trans-tidal waves, and are generated by
fluctuations in magnetism between the Earth's crust and atmosphere. The
magnetic push and pull between the Moon and the Sun creates waves with a
slightly shorter wavelength, known as tides. Their period and wavelength
can also range from a few hours to more than a day, and from a few
hundred to a thousand kilometers respectively. Storm surge tend to have
a slightly shorter wavelength and period compared to tides. When low
atmospheric pressure systems and high wind speeds in a storm it elevates
the ocean surface, generating storm surge, which may cause flooding in
coastal areas as it approaches the coastline. Tsunami's are on the lower
end of the scale, and are generated by earthquakes or submarine
`landslides. Their random nature makes them difficult to predict and
increase amplitude as they approach the coast. This can make them
considerably large along coastal areas and often cause immense damage
and loss of life.

\hypertarget{measuring-waves}{%
\subsection{Measuring waves}\label{measuring-waves}}

Waves are often thought of as an elevation of the sea surface from a
specific point over a period of time but this is obviously not he case.
This is known as \emph{surface elevation} and is the instantaneous
elevation of the sea surface above a specific point in a time record
(see figure \ref{fig:surface elevation figures}). Although surface
elevation does not represent a wave, it can be used to create a wave
profile. This is achieved by profiling the the surface elevation between
two successive \emph{downward zero-crossings} or \emph{upwardward
zero-crossings} (see figure \ref{fig:surface elevation figures}). Both
zero-crossings are symmetrical and essentially the same statistically.
However, in practice the downward zero-crossings are preferred as it
takes steepness of a approaching wave into account (the front, see
figure \ref{fig:surface elevation figures}) which is relevant to
characterising breaking waves. It is important to note that surface
elevation can be negative while a wave profile cannot.

\begin{figure}

{\centering \includegraphics{thesis_chapter_1_files/figure-latex/surface elevation figures-1}

}

\caption{\label{fig:surface elevation figures}The definition of a wave in a time record of the surface elevation with downward zero-crossings (upper panel) and upward zero-crossings (lower panel).}\label{fig:surface elevation figures}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics{thesis_chapter_1_files/figure-latex/Hs_Tp_surface_elevation figure-1}

}

\caption{The definition of wave height and wave period in a time record of the surface elevation (the wave is defined with downwrd zero-crossings.}\label{fig:Hs_Tp_surface_elevation figure}
\end{figure}

Waves are complex and various approaches, techniques and devices have
been developed in order to measure waves effectively. One of the ways
that has been used extensively used in the past but less so today, are
visual estimates which can be used to characterise \emph{significant
swell height} (Hs) and \emph{significant swell period} (Tp). Visual
estimates use 15 - 20 of the most well defined, higher waves of a number
of wave groups to characterise Hs and Tp. Although these parameters are
useful, they do not accurately reflect the waves that are occurring in
nature. Ocean waves are a combination of wind sea (short, irregular,
locally generated waves) and swell (long, smooth waves, generated by
distant storms) and so more parameters are needed to separate Hs and Tp
driven by wind sea and swell, i.e.~separate Hs and Tp parameters for
wind sea and swell. However, even with separate parameters for the
different types of waves would still not be enough to effectively
characterise waves in the ocean. In order to characterise the detail and
complexity of ocean waves a different approach must be used, know as the
\emph{spectral} approach. This approach is based on the idea that the
sea surface can be characterised as the summation of a large number of
harmonic wave components, calculated by the
\emph{random-phase/amplitude} model. The random-phase amplitude model is
a summation of wave components on a discrete scale, however a continuous
scale is more relevant to characterising and measuring of waves in the
ocean. The discrete spectrums are converted to continuous spectrums
through various arithmetic methods. Therefore, the random-phase
amplitude model is used to obtain a \emph{continuous variance density
spectrum} which is a statistical relevant measure for scientists and
engineers.
